---
title: "exam_2017_mysol"
author: "David Björelind"
date: "10/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Include packages here
library(bnlearn)
library(gRain)
library(HMM)

```

## 1. Graphical Models
```{r}

### a) ###
data("asia")
train = x
# Using HC to learn structure and parameters
network = hc(asia, restart = 10)
plot(network)
parameters = bn.fit(network, asia) # Conditional parameters for all nodes

grain = as.grain(parameters) # Conditional probabilities, full compared to network (bn.fit) Moralize and triangulate
structure = compile(grain) # creating junction tree, separators & residuals. Potentials for the cliques
# Want to test D _|_ X | E

evi = setEvidence(structure, nodes = c("X", "E"), states = c("no", "no")) 
dist1 = querygrain(evi, nodes = c("D"))
evi = setEvidence(structure, nodes = c("X", "E"), states = c("no", "yes")) 
dist2 = querygrain(evi, nodes = c("D"))
evi = setEvidence(structure, nodes = c("X", "E"), states = c("yes", "no")) 
dist3 = querygrain(evi, nodes = c("D"))
evi = setEvidence(structure, nodes = c("X", "E"), states = c("yes", "yes")) 
dist4 = querygrain(evi, nodes = c("D"))

### b) ###
# Essential DAGs
network = hc(asia, restart = 10)
plot(network)
a = cpdag(network)
plot(a)

# Generating graphs
graphs = random.graph(c("A","B","C","D","E"), num = 50000)
graphs = unique(graphs)
cpdags = lapply(graphs, cpdag)

# Comparing graphs
count = 0
for (i in 1:length(cpdags)){
  if(all.equal(graphs[[i]], cpdags[[i]]) == TRUE){
    count = count+1
  }
}

# Ratio
ratio = count/length(graphs)

### Essential DAG: ###
# Not Markov equivalent to any other DAG. Aka the union of all Markov equivalent DAGs

### Markov equivalent: ###
# Have the same skeleton
# Having the same colliders
```

## 2. Hidden Markov Models
```{r}
## Part a)
# Building a Hidden Markov Model
state = rep(1:100) # Actual number of states that the robot can be in (Hidden) Z nodes
probs = rep(1/100, 100)
symbols = 1:2 # States that we can observe (Not hidden) ### door or not? X nodes

funcmod = function(i){
  if(i == 100){
    return(i)
  } else if(i<=0){
    return(i+100)
  }else{
    return(i%%100)
  }
}

# If robot is in sector i:
transP = matrix(0, nrow = length(state), ncol = length(state))
for(i in 1:length(state)){
  transP[i,i] = 0.1
  transP[i, funcmod(i+1)] = 0.9
}

emissionP = matrix(.1, length(state), length(symbols))
emissionP[,2] = 0.9
emissionP[c(10,11,12,20,21,22,30,21,32),] = c(0.9,0.1)

# Initializing hidden markov model
robot = initHMM(States = state, Symbols = symbols, startProbs = probs, transProbs = transP, emissionProbs = emissionP)

## Part b)
which.maxima = function(x){
  return(which(x==max(x)))
}

obs = c(1,1,1,1,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2) # Robot at 3rd door and then moves into nothingness
post = posterior(robot, obs)

apply(post, 2, which.maxima)
```

## 3. Gaussian Processes
```{r}
library(mvtnorm)
library(kernlab)
# Covariance function
SquaredExpKernel <- function(x1,x2,sigmaF=1,l=3){
  n1 <- length(x1)
  n2 <- length(x2)
  K <- matrix(NA,n1,n2)
  for (i in 1:n2){
    K[,i] <- sigmaF^2*exp(-0.5*( (x1-x2[i])/l)^2 )
  }
  return(K)
}
# Returning a kernel: Squared Exponential
SEKernelfunc = function(ell, sigmaf){
  kernel = function(x1, x2){
  r = sqrt(sum((x1 - x2)^2))
  return(sigmaf^2*exp(-r^2/(2*ell^2)))
  }
  class(kernel) <- "kernel"
  return(kernel)
}

sigmaF = sqrt(1)
ell = c(0.2,1)
xGrid = seq(-1,1, by=0.1)

# Prior distribution of f (ell=0.2)
k = SEKernelfunc(ell[1], sigmaF)
k_matrix = kernelMatrix(k, xGrid, xGrid)
prior = rmvnorm(n=1, mean = rep(0,length(xGrid)) , sigma = k_matrix)
plot(x=xGrid, y=prior, type='l', main="ell = 0.2")
cols = c("blue", "red", "green", "orange")
for(i in 1:4){
  prior = rmvnorm(n=1, mean =rep(0,length(xGrid)) , sigma = k_matrix)
  lines(x=xGrid, y=prior, type='l', col = cols[i])
}

# Prior distribution of f (ell=1)
k = SEKernelfunc(ell[2], sigmaF)
k_matrix = kernelMatrix(k, xGrid, xGrid)
prior = rmvnorm(n=1, mean = rep(0,length(xGrid)) , sigma = k_matrix)
plot(x=xGrid, y=prior, type='l', main="ell = 1")
cols = c("blue", "red", "green", "orange")
for(i in 1:4){
  prior = rmvnorm(n=1, mean =rep(0,length(xGrid)) , sigma = k_matrix)
  lines(x=xGrid, y=prior, type='l', col = cols[i])
}
```
Ell = 1 produces much smoother priors!! Produces very different result when run multiple times.

## (i) and (ii)
```{r}
# ell = 0.2
k = SEKernelfunc(ell[1], sigmaF)
k(0,0.1)
k(0,0.5)

# ell = 1
k = SEKernelfunc(ell[2], sigmaF)
k(0,0.1)
k(0,0.5)
```
For a larger **ell**, it takes a larger difference between x1 and x2 for the correlation funciton to decrease (0.04 vs. 0.88).
Note here that correlation = covariance because sigmaF = 1.

## b)
```{r}
load("GPdata.RData")
sigmaN = 0.2
sigmaF = 1
ell = c(0.2, 1)
y_mean = mean(y)
y_sd = sd(y)
n <- length(x)
xgrid = seq(min(x), max(x), length.out = n)

# i) Posterior mean
kernelFunc = SEKernelfunc(ell[1],sigmaF)
GPfit = gausspr(x=scale(x), y=scale(y), kernel = kernelFunc, var = sigmaN, type="regression")
#meanpred = predict(GPfit, scale(x))
meanpred = predict(GPfit, scale(xgrid))

plot(x = xgrid, y = meanpred*y_sd+y_mean, col="red", type='l', ylab="y", main="x vs. y")

  ### 95% probability interval for f (This is done in the lab!!)

# Covariance matrix
Kss <- kernelMatrix(kernel = kernelFunc, x = xgrid, y = xgrid)
Kxx <- kernelMatrix(kernel = kernelFunc, x = x, y = x)
Kxs <- kernelMatrix(kernel = kernelFunc, x = x, y = xgrid)
Covf = Kss-t(Kxs)%*%solve(Kxx + sigmaN^2*diag(n), Kxs)

plot(x = xgrid, y = meanpred*y_sd+y_mean, col="red", type='l', lwd = 2, ylab="y", main="ell = 0.2")
lines(x = x, y = y, col="black", type='p', cex = 0.5)
lines(x = xgrid, y = meanpred*y_sd+y_mean + 1.96*sqrt(diag(Covf)), col="blue", type='l')
lines(x = xgrid, y = meanpred*y_sd+y_mean - 1.96*sqrt(diag(Covf)), col="blue", type='l')

### 95% prediction interval for a new data point y
lines(x = xgrid, y = meanpred*y_sd+y_mean + 1.96*sqrt(diag(Covf) + sigmaN^2), col="green", type='l')
lines(x = xgrid, y = meanpred*y_sd+y_mean - 1.96*sqrt(diag(Covf) + sigmaN^2), col="green", type='l')

### ell = 1
kernelFunc = SEKernelfunc(ell[2],sigmaF)
GPfit = gausspr(x=scale(x), y=scale(y), kernel = kernelFunc, var = sigmaN, type="regression")
meanpred = predict(GPfit, scale(xgrid))

n <- length(x)
xgrid = seq(min(x), max(x), length.out = n)
Kss <- kernelMatrix(kernel = kernelFunc, x = xgrid, y = xgrid)
Kxx <- kernelMatrix(kernel = kernelFunc, x = x, y = x)
Kxs <- kernelMatrix(kernel = kernelFunc, x = x, y = xgrid)
Covf = Kss-t(Kxs)%*%solve(Kxx + sigmaN^2*diag(n), Kxs)

plot(x = xgrid, y = meanpred*y_sd+y_mean, col="red", type='l', lwd = 2, ylab="logRatio", main="ell = 1")
lines(x = x, y = y, col="black", type='p', cex = 0.5)
lines(x = xgrid, y = meanpred*y_sd+y_mean + 1.96*sqrt(diag(Covf)), col="blue", type='l')
lines(x = xgrid, y = meanpred*y_sd+y_mean - 1.96*sqrt(diag(Covf)), col="blue", type='l')

### 95% prediction interval for a new data point y
lines(x = xgrid, y = meanpred*y_sd+y_mean + 1.96*sqrt(diag(Covf) + sigmaN^2), col="green", type='l')
lines(x = xgrid, y = meanpred*y_sd+y_mean - 1.96*sqrt(diag(Covf) + sigmaN^2), col="green", type='l')

```
A larger **ell** makes to produced means smoother. The different between (ii) and (iii) is that we add the noise from the data. To take this into consideration, the intervals becomes wider.

A problem with the kernel, Squared Exponential, is that the data has different smoothness for small x and large x.

GIVEN ANSWERS:
# Question: Explain the difference between the results from ii) and iii). 
# Answer: ii) is about the uncertainty of the function f, which is the MEAN of y
#         iii) is about the uncertainty of individual y values. They are uncertain for 
#              two reasons: you don't know f at the test point, and you don't know
#              the error (epsilon) that will hit this individual observation

# Question: Discuss the differences in results from using the two length scales.
#           Answer: shorter length scale gives less smooth f. We are overfitting the data.
#           Answer: longer length scale gives more smoothness.

# Question: Do you think a GP with a squared exponential kernel is a good model for this data? If not, why?
#           Answer: One would have to experiment with other length scales, or estimate
#           the length scales (see question 3c), but this is not likely to help here.
#           The issue is that the data seems to be have different smoothness for small x
#           than it has for large x (where the function seems much more flat)
#           The solution is probably to have different length scales for different x


### Question 3(c)

# For full points here you should mention EITHER of the following two approaches:
# 1. The marginal likelihood can be used to select optimal hyperparameters, 
# and also the noise variance. We can optimize the log marginal likelihood with
# respect to the hyperparameters. In Gaussian Process Regression the marginal likelihood
# is availble in closed form (a formula).
# 2. We can use sampling methods (e.g. MCMC) to sample from the marginal posterior of the hyperparameters
# We need a prior p(theta) for the hyperparameter and then Bayes rule gives the marginal posterior
# p(theta | data) propto p(data | theta)*p(theta)
# where p(data | theta) is the marginal likelihood (f has been integrated out).

# If the noise variance is unknown, we can treat like any of the kernel hyperparameters and infer the noise variance 
# jointly with the length scale and the prior variance sigma_f
