---
title: "Lab 4 - Gaussian Processes"
author: "David Björelind"
date: "10/18/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Include packages here
library(kernlab)
library(AtmRay)

# Covariance function
SquaredExpKernel <- function(x1,x2,sigmaF=1,l=3){
  n1 <- length(x1)
  n2 <- length(x2)
  K <- matrix(NA,n1,n2)
  for (i in 1:n2){
    K[,i] <- sigmaF^2*exp(-0.5*( (x1-x2[i])/l)^2 )
  }
  return(K)
}
```


## 2.1) (1) Implementing GP Regression
```{r}
# Simulating from posterior distribution of f
posteriorGP = function(X, y, XStar, sigmaNoise, k, sigmaF, l){
  
  K = k(X,X, sigmaF, l)
  n = length(XStar)
  L = t(chol(K + sigmaNoise*diag(dim(K)[1])))
  kStar = k(X,XStar, sigmaF, l)
  alpha = solve(t(L), solve(L,y))
  
  FStar = t(kStar) %*% alpha
  v = solve(L, kStar)
  vf = k(XStar, XStar, sigmaF, l) - t(v)%*%v #+ sigmaNoise*diag(n) #Adding sigma for noise
  #print(k(XStar, XStar, sigmaF, l))
  #print(diag(t(v)%*%v))
  logmarglike = -t(y)%*%alpha/2 - sum(diag(L)) - n/2*log(2*pi)
  
  
  # Returns a vector with the posterior mean and variance
  return(list("mean" = FStar,"variance" = vf,"logmarglike" = logmarglike))
}
```

## 2.1) (2) GP Regression with kernlab
```{r}
sigmaf = 1^2
sigman = 0.1^2
l = 0.3
x = 0.4
y = 0.719
xrange = seq(-1,1, length=25)

res = posteriorGP(x, y, xrange, sigman, SquaredExpKernel, sigmaF = sigmaf, l=l)

# (2) Plotting posterior mean and 95% interval bands
plot(x = xrange, y = res$mean, ylim = c(-2,2), main = "Mean for (x = 0.4, y = 0.719)")
lines(x = xrange, y = res$mean + sqrt(diag(res$variance))*1.96, col="blue", type="l")
lines(x = xrange, y = res$mean - sqrt(diag(res$variance))*1.96, col="blue",  type="l")

```

## 2.1) (3)
```{r}
# (3)
x = c(0.4,-0.6)
y = c(0.719,-0.044)

res = posteriorGP(x, y, xrange, sigman, SquaredExpKernel, sigmaF = sigmaf, l=l)
plot(x = xrange, y = res$mean, ylim = c(-2,2), main = "Updating mean for 2 observations")
lines(x = xrange, y = res$mean + sqrt(diag(res$variance))*1.96, col="blue", type="l")
lines(x = xrange, y = res$mean - sqrt(diag(res$variance))*1.96, col="blue",  type="l")
```

## 2.1) (4)
```{r}
# (4)
x = c(-1.0, -0.6, -0.2, 0.4, 0.8) 
y = c(0.768, -0.044, -0.940, 0.719, -0.664)
sigmaf = 1^2
l = 0.3
res = posteriorGP(x, y, xrange, sigman, SquaredExpKernel, sigmaF = sigmaf, l=l)
plot(x = xrange, y = res$mean, ylim = c(-2,2), main = "Updating mean with 5 observations")
lines(x = xrange, y = res$mean + sqrt(diag(res$variance))*1.96, col="blue", type="l")
lines(x = xrange, y = res$mean - sqrt(diag(res$variance))*1.96, col="blue",  type="l")
```
## 2.1) (5)
```{r}
# (5)
x = c(-1.0, -0.6, -0.2, 0.4, 0.8) 
y = c(0.768, -0.044, -0.940, 0.719, -0.664)
sigmaf = 1
l = 1
res = posteriorGP(x, y, xrange, sigman, SquaredExpKernel, sigmaF = sigmaf, l=l)

plot(x = xrange, y = res$mean, ylim = c(-3,2), main = "Updating mean with 5 observations and l=1")
lines(x = xrange, y = res$mean + sqrt(diag(res$variance))*1.96, col="blue", type="l")
lines(x = xrange, y = res$mean - sqrt(diag(res$variance))*1.96, col="blue",  type="l")
```
Since **l=1**, the produced function will be more smooth compared to (4). We also see that the bands created are much more narrow compared to plot produced in (4).


## 2.2 (1) GP Regression with kernlab
```{r, include=FALSE}
temps = read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/TempTullinge.csv", header=TRUE, sep=";")$temp
time = seq(from=1, to=2190, by=5)
temps = temps[time]
day = rep(seq(from=1, to=361, by=5), times=6)
```

## 2.2 (1)
```{r}

SEKernel = function(x1, x2, ell = 1, sigmaf = 1){
  r = sqrt(sum((x1 - x2)^2))
  return(sigmaf^2*exp(-r^2/(2*ell^2)))
}

kernel = SEKernel(x1=1, x2=2, ell = l, sigmaf = sigmaf)

X = c(1,3,4)
XStar = c(2,3,4)
covarmatrix = kernelMatrix(kernel = SEKernel, x=X, y=XStar)

newKernelMatrix = function(kernel, x, y){
  SEKernel = function(x1, x2, ell = 1, sigmaf = 1){
    r = sqrt(sum((x1 - x2)^2))
    return(sigmaf*exp(-r^2/(2*ell^2)))
  }
  return(covarmatrix = kernelMatrix(kernel = SEKernel, x=x, y=y))
}
SEKernelfunc = function(ell, sigmaf){
  
  kernel = function(x1, x2){
  r = sqrt(sum((x1 - x2)^2))
  return(sigmaf^2*exp(-r^2/(2*ell^2)))
  }
  class(kernel) <- "kernel"
  return(kernel)
}

```


##2.2 (2)
```{r}
ell = 0.2
sigmaf = 20

# Getting the error term for the GP
polyFit <- lm(temps ~  time + I(time^2))
sigma = sd(polyFit$residuals)
GPfit = gausspr(x=time, y=temps, kernel = SEKernelfunc(ell, sigmaf), var = sigma^2, type="regression")
meanpred = predict(GPfit, time)

# Plotting the means!
plot(time, meanpred, col="red", type='l')

```
A higher value for ell increases smoothness.
A higher value for sigmaf increases the possible ranges of the predicted means. A lower value gives lower covariance between prediction. This explains the difference in the highest and lowest values for different sigmaf's.

##2.2 (3)
```{r}
# Modified posteriorGP function
posteriorGP = function(X, y, XStar, sigmaNoise, k){
  K = kernelMatrix(k,X,X)
  n = length(XStar)
  L = t(chol(K + sigmaNoise*diag(dim(K)[1])))
  kStar = kernelMatrix(k,X,XStar)
  alpha = solve(t(L), solve(L,y))
  FStar = (t(kStar)) %*% alpha
  v = solve(L, kStar)
  vf = kernelMatrix(k,XStar, XStar) - t(v)%*%v #+ sigmaNoise*diag(n) #Adding sigma for noise
  logmarglike = -t(y)%*%alpha/2 - sum(diag(L)) - n/2*log(2*pi)
  
  # Returns a vector with the posterior mean and variance
  return(list("mean" = FStar,"variance" = vf,"logmarglike" = logmarglike))
}

var = posteriorGP(X=scale(time), y=scale(temps), XStar=scale(time), sigmaNoise=sigma^2, k=SEKernelfunc(ell, sigmaf))
varr = sqrt(var$variance)*sd(temps)

plot(time, meanpred, col="red", type='l', ylim=c(-25,25), main="Plot with means from time and 95% interval")
lines(x = time, y = meanpred + sqrt(diag(var$variance)*1.96*sd(temps)), col="blue", type="l")
lines(x = time, y = meanpred - sqrt(diag(var$variance)*1.96*sd(temps)), col="blue",  type="l")
lines(x = time, y = temps,  type="p")
```

##2.2 (4)
```{r}
ell = 0.2
sigmaf = 20

# Getting the error term for the GP, using day instead of time
polyFit <- lm(temps ~  day + I(day^2))
sigma = sd(polyFit$residuals)
GPfit = gausspr(x=day, y=temps, kernel = SEKernelfunc(ell, sigmaf), var = sigma^2, type="regression")
meanpred_day = predict(GPfit, day)

plot(time, meanpred, col="red", type='l', ylim=c(-25,30), main="Plot with predictions from day (red:time, green:day)")
lines(x = time, y = meanpred_day + sqrt(diag(var$variance)*1.96*sd(temps)), col="blue", type="l")
lines(x = time, y = meanpred_day - sqrt(diag(var$variance)*1.96*sd(temps)), col="blue",  type="l")
lines(x = time, y = temps,  type="p")
lines(x = time, y = meanpred_day, col="green",  type="l")

```

##2.2 (5)
```{r}
periodickernelfunc = function(sigmaf, l1, l2, d){
  periodicKernel = function(x1, x2){
  r = sqrt(sum((x1 - x2)^2))
  one = exp(-2*sin(r/d)/l1^2)
  two = exp(-r^2/(2*l2^2))
  return(sigmaf^2*one*two)
  }
  class(periodicKernel) <- "kernel"
  return(periodicKernel)
}

# Hyperparameters
sigmaf = 20
l1 = 1
l2 = 10
d = 365/sd(time)

# Getting the error term for the GP # USING TIME MAYBE CHANGE
polyFit <- lm(temps ~  time + I(time^2))
sigma = sd(polyFit$residuals)
GPfit = gausspr(x=time, y=temps, kernel = periodickernelfunc(sigmaf, l1, l2, d), var = sigma^2, type="regression")
meanpred_5 = predict(GPfit, time)

# Plotting the means!
plot(x = time, y = meanpred_5, col="red", type='l', ylim=c(-25,25),main = "Means using the periodic kernel")
lines(x = time, y = meanpred_5 + sqrt(diag(var$variance)*1.96*sd(temps)), col="blue", type="l")
lines(x = time, y = meanpred_5 - sqrt(diag(var$variance)*1.96*sd(temps)), col="blue",  type="l")
lines(x = time, y = temps,  type="p")
```
We can see that the periodic kernel produces produces more sharp "valleys" and more even "tops" compared to the more smooth Squared Exponential kernel.

##2.3 (1)
```{r}
#data <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/banknoteFraud.csv", header=FALSE, sep=",") names(data) <- c("varWave","skewWave","kurtWave","entropyWave","fraud") data[,5] <- as.factor(data[,5])

```

## Another chunk
```{r}
# (3)

```

## Another chunk
```{r}
# (3)

```